---
title: "STAT 452/652 Final Project Code"
author: "Shivansh Ghai"
format: html
date: "December 4, 2025"
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# -------------------------------------------------------

# BLOCK 1: Load data + basic structure checks

# -------------------------------------------------------

```{r}
# 1. Load training + test predictor files
train <- read.csv("training_data.csv")
test  <- read.csv("test_predictors.csv")

# 2. Basic structure
cat("=== TRAIN STRUCTURE ===\n")
str(train)

cat("\n=== TEST STRUCTURE ===\n")
str(test)

# 3. Missing value check
cat("\n=== MISSING VALUES (train) ===\n")
print(colSums(is.na(train)))

cat("\n=== MISSING VALUES (test) ===\n")
print(colSums(is.na(test)))

# 4. Summary statistics (quick view)
cat("\n=== SUMMARY(train) ===\n")
print(summary(train))

# 5. Check that predictor names match
cat("\n=== MATCHING PREDICTOR NAMES? ===\n")
print(setequal(colnames(train)[-1], colnames(test)))
```

# -------------------------------------------------------

# BLOCK 2: CV setup + baseline OLS regression

# -------------------------------------------------------

```{r}
set.seed(452)

# Libraries
library(caret)
library(glmnet)
library(randomForest)
library(gbm)

# 1. Separate X and y
X <- train[, grep("^X", names(train))]  # predictors only
y <- train$Y                            # response

# 2. Define cross-validation
ctrl <- trainControl(
  method = "cv",
  number = 10
)

cat("=== Cross-validation setup complete ===\n\n")

# 3. Fit baseline linear regression model
cat("=== Fitting OLS model (no regularization) ===\n")
ols_fit <- train(
  x = X,
  y = y,
  method = "lm",
  trControl = ctrl
)

print(ols_fit)

# Display CV RMSE
cat("\n=== OLS Cross-Validated RMSE ===\n")
print(ols_fit$results$RMSE)
```

# -------------------------------------------------------

# BLOCK 3: Ridge Regression (alpha = 0)

# -------------------------------------------------------

```{r}
set.seed(452)

# Grid of lambda values on a log scale
ridge_grid <- expand.grid(
  alpha = 0,
  lambda = 10^seq(-3, 3, length = 50)
)

cat("=== Fitting Ridge Regression ===\n")

ridge_fit <- train(
  x = X,
  y = y,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = ridge_grid
)

# Print model summary
print(ridge_fit)

# Best tuning parameter
cat("\n=== Best Ridge Lambda ===\n")
print(ridge_fit$bestTune)

# Cross-validated RMSE
cat("\n=== Ridge CV RMSE (best) ===\n")
best_rmse_ridge <- ridge_fit$results$RMSE[
  ridge_fit$results$lambda 
  == ridge_fit$bestTune$lambda]
print(best_rmse_ridge)
```

# -------------------------------------------------------

# BLOCK 4: Lasso Regression (alpha = 1)

# -------------------------------------------------------

```{r}
set.seed(452)

# Lasso grid of lambdas
lasso_grid <- expand.grid(
  alpha = 1,
  lambda = 10^seq(-3, 3, length = 50)
)

cat("=== Fitting LASSO Regression ===\n")

lasso_fit <- train(
  x = X,
  y = y,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = lasso_grid
)

print(lasso_fit)

# Best lambda
cat("\n=== Best Lasso Lambda ===\n")
print(lasso_fit$bestTune)

# Best CV RMSE
cat("\n=== Lasso CV RMSE (best) ===\n")
best_rmse_lasso <- lasso_fit$results$RMSE[
  lasso_fit$results$lambda == lasso_fit$bestTune$lambda
]
print(best_rmse_lasso)

# Extract non-zero coefficients
cat("\n=== Number of Non-Zero Coefficients ===\n")
lasso_coefs <- coef(lasso_fit$finalModel, s 
                    = lasso_fit$bestTune$lambda)
print(sum(lasso_coefs != 0))   # includes intercept
```

# -------------------------------------------------------

# BLOCK 5: Elastic Net Regression (α tuning)

# -------------------------------------------------------

```{r}
set.seed(452)

library(glmnet)

X_mat <- as.matrix(train[, -which(names(train) 
                                  == "Y")])
y_vec <- train$Y

# Candidate alpha values (mix between ridge & lasso)
alpha_grid <- seq(0.1, 0.9, by = 0.2)

en_results <- list()
en_best <- list()
en_mspe <- c()

for (a in alpha_grid) {
  cv_fit <- cv.glmnet(
    x = X_mat,
    y = y_vec,
    alpha = a,
    nfolds = 10
  )
  en_results[[as.character(a)]] <- cv_fit
  en_mspe <- c(en_mspe, cv_fit$cvm[cv_fit$lambda 
                                   == cv_fit$lambda.min])
}

names(en_mspe) <- paste0("alpha=", alpha_grid)

cat("\n=== Elastic Net CV Errors ===\n")
print(en_mspe)

best_alpha <- alpha_grid[which.min(en_mspe)]
best_alpha
```

# -------------------------------------------------------

# BLOCK 6: Principal Components Regression (PCR)

# -------------------------------------------------------

```{r}
set.seed(452)

library(pls)

pcr_fit <- pcr(
  Y ~ .,
  data = train,
  scale = TRUE,
  validation = "CV"
)

cat("\n=== PCR Cross-Validation Summary ===\n")
summary(pcr_fit)

# Select #components using CV
validationplot(pcr_fit, val.type = "MSEP")
```

# -------------------------------------------------------

# BLOCK 7: Partial Least Squares Regression (PLS)

# -------------------------------------------------------

```{r}
set.seed(452)

pls_fit <- plsr(
  Y ~ .,
  data = train,
  scale = TRUE,
  validation = "CV"
)

cat("\n=== PLS Cross-Validation Summary ===\n")
summary(pls_fit)

validationplot(pls_fit, val.type = "MSEP")
```

# -------------------------------------------------------

# BLOCK 8: Natural Spline Model (df = 4 each predictor)

# -------------------------------------------------------

```{r}
library(splines)

set.seed(452)

ns_formula <- as.formula(
  paste("Y ~", paste(paste0("ns(", names(train)[names(train)!="Y"], ", df=4)"), 
                     collapse=" + "))
)

ns_fit <- lm(ns_formula, data = train)

cat("\n=== Natural Splines Model Summary ===\n")
print(summary(ns_fit))
```

# -------------------------------------------------------

# BLOCK 9: Smoothing Spline Model (via GAM)

# -------------------------------------------------------

```{r}
set.seed(452)
library(gam)

smooth_formula <- as.formula(
  paste("Y ~", paste(paste0("s(", 
                            names(train)[names(train)!=
                                           "Y"], ")"), collapse=" + ")))
ss_fit <- gam(smooth_formula, data = train)
cat("\n=== Smoothing Spline GAM Summary ===\n")
print(summary(ss_fit))
cat("\n=== Smoothing Spline GAM Summary ===\n")
print(summary(ss_fit))
```

# -------------------------------------------------------

# BLOCK 10: Regression Tree (rpart)

# -------------------------------------------------------

```{r}
library(rpart)
library(rpart.plot)

set.seed(452)

tree_fit <- rpart(
  Y ~ .,
  data = train,
  method = "anova",
  cp = 0.001
)

cat("\n=== Regression Tree Summary ===\n")
printcp(tree_fit)

rpart.plot(tree_fit)
```

# -------------------------------------------------------

# BLOCK 11: Random Forest

# -------------------------------------------------------

```{r}
set.seed(452)

rf_grid <- expand.grid(
  mtry = c(3, 5, 7, 9, 11)   # try varying # of variables per split
)

cat("=== Fitting Random Forest ===\n")

rf_fit <- train(
  x = X,
  y = y,
  method = "rf",
  trControl = ctrl,
  tuneGrid = rf_grid,
  importance = TRUE,
  ntree = 500   # solid default
)

print(rf_fit)

# Best mtry
cat("\n=== Best mtry ===\n")
print(rf_fit$bestTune)

# Best RMSE
cat("\n=== RF CV RMSE (best) ===\n")
best_rmse_rf <- rf_fit$results$RMSE[
  rf_fit$results$mtry == rf_fit$bestTune$mtry]
print(best_rmse_rf)

# Variable importance
imp <- varImp(rf_fit)

cat("=== Variable Importance Structure ===\n")
print(str(imp))

# Extract the importance matrix correctly
imp_mat <- imp$importance

# Order by importance
ordered <- imp_mat[order(-imp_mat[ ,1]), , drop = FALSE]

cat("\n=== TOP 10 MOST IMPORTANT VARIABLES ===\n")
print(ordered[1:10, , drop = FALSE])
```

# -------------------------------------------------------

# BLOCK 12: Gradient Boosting Machine (GBM)

# -------------------------------------------------------

```{r}
set.seed(452)

gbm_grid <- expand.grid(
  n.trees = c(300, 600, 1000),
  interaction.depth = c(1, 3, 5),
  shrinkage = c(0.01, 0.05),
  n.minobsinnode = 10
)

cat("=== Fitting Gradient Boosting Model (GBM) ===\n")

gbm_fit <- train(
  x = X,
  y = y,
  method = "gbm",
  trControl = ctrl,
  verbose = FALSE,
  tuneGrid = gbm_grid
)

print(gbm_fit)

# Best parameters
cat("\n=== Best GBM Parameters ===\n")
print(gbm_fit$bestTune)

# Best RMSE
cat("\n=== GBM CV RMSE (best) ===\n")
best_rmse_gbm <- gbm_fit$results$RMSE[
  gbm_fit$results$n.trees == gbm_fit$bestTune$n.trees &
    gbm_fit$results$interaction.depth ==
    gbm_fit$bestTune$interaction.depth &
    gbm_fit$results$shrinkage == 
    gbm_fit$bestTune$shrinkage
]
print(best_rmse_gbm)
```

# -------------------------------------------------------

# BLOCK 13: Tuned Regression Tree (cp via 1-SE rule)

# -------------------------------------------------------

```{r}
set.seed(452)

tree_base <- rpart(
  Y ~ .,
  data   = train,
  method = "anova",
  cp     = 0.001
)

cat("\n=== Base Regression Tree CP Table ===\n")
cp_table <- tree_base$cptable   # <-- FIXED
print(cp_table)

# Minimum xerror row
min_row  <- which.min(cp_table[, "xerror"])
xerr_min <- cp_table[min_row, "xerror"]
xerr_se  <- cp_table[min_row, "xstd"]

# 1-SE rule: smallest tree whose xerror <= min + 1 SE
cp_1se <- cp_table[cp_table[, "xerror"] <= xerr_min 
                   + xerr_se, ][1, "CP"]

cat("\nChosen cp (1-SE rule):", cp_1se, "\n")

tree_tuned <- prune(tree_base, cp = cp_1se)

# Training MSPE
tree_pred_train <- predict(tree_tuned, newdata = train)
tree_mspe_train <- mean((train$Y - tree_pred_train)^2)

cat("\n=== Tuned Regression Tree Training MSPE ===\n")
print(tree_mspe_train)
```

# -------------------------------------------------------

# BLOCK 14: Tuned Random Forest (mtry × nodesize grid)

# -------------------------------------------------------

```{r}
set.seed(452)
library(randomForest)

X <- train[, -which(names(train) == "Y")]
y <- train$Y

p <- ncol(X)

mtry_grid     <- c(round(sqrt(p)), 11, round(p/2))  
# e.g. 4–5, 11, 9–10
nodesize_grid <- c(2, 5, 10)

rf_grid <- expand.grid(
  mtry     = mtry_grid,
  nodesize = nodesize_grid
)

rf_results <- data.frame(
  mtry     = integer(),
  nodesize = integer(),
  OOB_MSE  = double()
)

for (i in seq_len(nrow(rf_grid))) {
  m  <- rf_grid$mtry[i]
  ns <- rf_grid$nodesize[i]
  
  rf_fit <- randomForest(
    x         = X,
    y         = y,
    mtry      = m,
    nodesize  = ns,
    ntree     = 500,
    importance = FALSE
  )
  
  oob_mse <- tail(rf_fit$mse, 1)  # last OOB MSE
  
  rf_results <- rbind(
    rf_results,
    data.frame(mtry = m, nodesize = ns, OOB_MSE = oob_mse)
  )
  
  cat("Finished RF combo:", "mtry =", m, ", 
      nodesize =", ns,
      ", OOB MSE =", round(oob_mse, 3), "\n")
}

cat("\n=== Random Forest Tuning Results ===\n")
print(rf_results)

best_idx          <- which.min(rf_results$OOB_MSE)
best_rf_settings  <- rf_results[best_idx, ]

cat("\n=== Best RF Settings ===\n")
print(best_rf_settings)

best_mtry     <- best_rf_settings$mtry
best_nodesize <- best_rf_settings$nodesize
```

# -------------------------------------------------------

# BLOCK 15: Fully Tuned Gradient Boosting (GBM grid)

# -------------------------------------------------------

```{r}
library(gbm)

set.seed(452)

gbm_grid <- expand.grid(
  interaction.depth = c(2, 3, 4),
  n.minobsinnode    = c(10, 30)
)

gbm_results <- data.frame(
  depth     = integer(),
  nmin      = integer(),
  best_iter = integer(),
  cv_mse    = double()
)

for (i in seq_len(nrow(gbm_grid))) {
  d   <- gbm_grid$interaction.depth[i]
  nmn <- gbm_grid$n.minobsinnode[i]
  
  cat("\n--- Fitting GBM combo", i,
      ": depth =", d, ", n.minobsinnode =", 
      nmn, " ---\n")
  
  gbm_fit <- gbm(
    formula           = Y ~ .,
    data              = train,
    distribution      = "gaussian",
    n.trees           = 3000,
    interaction.depth = d,
    shrinkage         = 0.01,
    n.minobsinnode    = nmn,
    bag.fraction      = 0.5,
    train.fraction    = 1.0,
    cv.folds          = 5,
    n.cores           = 1,
    verbose           = FALSE
  )
  
  best_iter <- gbm.perf(gbm_fit, method = "cv", 
                        plot.it = FALSE)
  best_mse  <- gbm_fit$cv.error[best_iter]
  
  gbm_results <- rbind(
    gbm_results,
    data.frame(
      depth     = d,
      nmin      = nmn,
      best_iter = best_iter,
      cv_mse    = best_mse
    )
  )
  
  cat("Best iter:", best_iter, " | CV MSE:", 
      round(best_mse, 3), "\n")
}

cat("\n=== GBM Tuning Results ===\n")
print(gbm_results)

best_gbm_idx      <- which.min(gbm_results$cv_mse)
best_gbm_settings <- gbm_results[best_gbm_idx, ]

cat("\n=== Best GBM Settings ===\n")
print(best_gbm_settings)
```

# -------------------------------------------------------

# Comparison

# -------------------------------------------------------

```{r}
model_perf <- data.frame(
  model  = c("OLS", "Ridge", "LASSO", 
             paste0("ElasticNet (alpha=", best_alpha, ")"),
             "RF (caret)", "RF (manual)", 
             "GBM (caret)", "GBM (manual)"),
  RMSE   = c(
    ols_fit$results$RMSE,
    best_rmse_ridge,
    best_rmse_lasso,
    sqrt(min(en_mspe)),                # ENet MSE → RMSE
    best_rmse_rf,
    sqrt(min(rf_results$OOB_MSE)),     # manual RF OOB RMSE
    best_rmse_gbm,
    sqrt(min(gbm_results$cv_mse))      # manual GBM RMSE
  )
)

cat("\n=== MODEL COMPARISON TABLE ===\n")
print(model_perf[order(model_perf$RMSE), ])
```

# -------------------------------------------------------

# FINAL MODEL SCRIPT

# Random Forest (mtry = 11, ntree = 500)

# -------------------------------------------------------

```{r}
set.seed(452)

# 1. Load libraries
library(randomForest)

# 2. Load training + test data
train <- read.csv("training_data.csv")
test  <- read.csv("test_predictors.csv")

# 3. Prepare X and y
X <- train[ , grep("^X", names(train))]
y <- train$Y

# 4. Fit the Final Random Forest on FULL training data
final_rf <- randomForest(
  x        = X,
  y        = y,
  mtry     = best_mtry,
  ntree    = 1500,      # see Refinement #2 below
  nodesize = best_nodesize,
  importance = TRUE
)

cat("=== Final Model Fit Complete ===\n")
print(final_rf)

# 5. Predict on the TEST set
final_predictions <- predict(final_rf, newdata = test)

# 6. Save predictions EXACTLY in required format
write.table(
  final_predictions,
  "my_predictions.csv",
  sep = ",",
  row.names = FALSE,
  col.names = FALSE
)

cat("\n=== Predictions Saved to my_predictions.csv ===\n")
```
